Exploratory data anlysis on white wines
========================================================

```{r echo=FALSE, message=FALSE, warning=FALSE, packages}
# Load all of the packages that you end up using in your analysis in this code
# chunk.

# Notice that the parameter "echo" was set to FALSE for this code chunk. This
# prevents the code from displaying in the knitted HTML output. You should set
# echo=FALSE for all code chunks in your file, unless it makes sense for your
# report to show the code that generated a particular plot.

# The other parameters for "message" and "warning" should also be set to FALSE
# for other code chunks once you have verified that each plot comes out as you
# want it to. This will clean up the flow of your report.

library(GGally)
library(ggplot2)
library(scales)
library(memisc)
library(gridExtra)
library(corrplot)
library(dplyr)
library(tidyverse)
```

```{r echo=FALSE, Load_the_Data}
# Load the Data
df <- read.csv('wineQualityWhites.csv')
```


# Dataset Section

## General information

The dataset *wineQualityWhites.csv* contains 4898 observations for 12 variables. These are: 

- fixed.acidity (tartaric acid - g / dm^3)
- volatile.acidity (acetic acid - g / dm^3)
- citric.acid (g / dm^3)
- residual.sugar (g / dm^3)
- chlorides (sodium chloride - g / dm^3
- free.sulfur.dioxide (mg / dm^3)
- total.sulfur.dioxide (mg / dm^3)
- density (g / cm^3)
- pH
- sulphates (potassium sulphate - g / dm3)
- alcohol (% by volume)
- quality (score between 0 and 10)

The dataset itself is dated back to 2009.

## Names in the dataset
```{r echo=FALSE, basics_1}
# get the names of each column in the dataframe
names(df)
```

## Dimension of the dataset
```{r echo=FALSE, basics_2}
# get the dimension of the dataframe
dim(df)
```

## Summary of the dataset
```{r echo=FALSE, basics_3}
# get a summary of the dataframe
summary(df)
```


# Univariate Plots Section

## fixed.acidity, volatile.acidity and citric.acid

```{r echo=FALSE, message=FALSE, Univariate_Plots_2}
# plotting a histogram for fixed.acidity
plt1 <- ggplot(aes(x = fixed.acidity), data = df) +
  geom_histogram()

# plotting a histogram for volatile.acidity
plt2 <- ggplot(aes(x = volatile.acidity), data = df) +
  geom_histogram()

# plotting a histogram for citric.acid
plt3 <- ggplot(aes(x = citric.acid), data = df) +
  geom_histogram()

# arranging all plots under each other (ncol = 1)
grid.arrange(plt1, plt2, plt3, ncol = 1)
```

By looking at fixed.acidity and volatile.acidity it seems like there comes a general acidity with every wine. The variability is neglible because the the basic acidity (fixed.acidity) is quite high. The same applies to citric.acid which is, when present, fairly low. 

## residual.sugar and chlorides

```{r echo=FALSE, message=FALSE, Univariate_Plots_3}
# plotting a histogram for residual.sugar 
plt4 <- ggplot(aes(x = residual.sugar), data = df) +
  geom_histogram()

# plotting a histogram for chlorides
plt5 <- ggplot(aes(x = chlorides), data = df) +
  geom_histogram()

# arranging the two plots under each other (ncol = 1)
grid.arrange(plt4, plt5, ncol = 1)
```

The plot for residual.sugar combined with the explanations for this dataset leeds to the conclusion that almost all wines are not sweet (a wine is sweet when residual.sugar is equal to or above 45).
The detected concentration of chlorides is quite low compared to sugar.

## free.sulfur.dioxide and total.sulfur.dioxide

``````{r echo=FALSE, message=FALSE, Univariate_Plots_4}
# plotting a histogram for free.sulfur.dioxide
plt6 <- ggplot(aes(x = free.sulfur.dioxide), data = df) +
  geom_histogram()

# plotting a histogram for total.sulfur.dioxide 
plt7 <- ggplot(aes(x = total.sulfur.dioxide), data = df) +
  geom_histogram()

# arranging the two plots under each other (ncol = 1)
grid.arrange(plt6, plt7, ncol = 1)
```

By looking at the distribution of both variables it becomes clear that the free.sulfur.dioxide has an influence on the total.sulfur.dioxide amount but bound forms of sulfur dioxide are playing an equal role.

```{r echo=FALSE, message=FALSE, Univariate_Plots_adding_variable }
# creating an additional variable called bound.sulfur.dioxide by substracting
# free.sulfur.dioxide from total.sulfur.dioxide
df$bound.sulfur.dioxide <- df$total.sulfur.dioxide - df$free.sulfur.dioxide
```

## density, pH, sulphates and alcohol

```{r echo=FALSE, message=FALSE, Univariate_Plots_5}
# plotting a histogram for density
plt8 <- ggplot(aes(x = density), data = df) +
  geom_histogram()

# plotting a histogram for pH
plt9 <- ggplot(aes(x = pH), data = df) +
  geom_histogram()

# plotting a histogram for sulphates
plt10 <- ggplot(aes(x = sulphates), data = df) +
  geom_histogram()

# plotting a histogram for alcohol
plt11 <- ggplot(aes(x = alcohol), data = df) +
  geom_histogram()

# arranging the four plots under each other (ncol = 1)
grid.arrange(plt8, plt9, plt10, plt11, ncol = 1)
```

With density concentrated below 1 one has to conclude that wine is not as dense as water.
The pH value describes how acidic a wine is. The observed values between 2.7 and 3.9 are normal (as per the provided explanation for the dataset). The wines tend to be more acidic.
The alcohol levels are relatively spread with a major peak between 8 and 10.

## quality

```{r echo=FALSE, message=FALSE, Univariate_Plots_6}
# plotting a histogram for alcohol with the median as a dotted white line
ggplot(aes(x = quality), data = df) +
  geom_histogram(binwidth = 1) +
  geom_vline(xintercept = mean(df$quality), color='white', linetype=2)
```

The wines were mostly of medium (6) quality.

## quality groups

For further analysis I categorized all wines into three groups:
- low: quality between 0 and 4
- medium: quality between 5 and 6
- high: quality between 7 and 10

```{r echo=FALSE, message=FALSE, Univariate_Plots_7}
# categorizing all wines into three groups (low: <= 4, medium: 5 <= x <= 6, high: >6)
df$quality.group <- cut(df$quality, breaks = c(0,4,6,10), 
                        labels=c('low','medium','high'))

#getting a summary
summary(df$quality.group)
```

```{r echo=FALSE, message=FALSE, Univariate_Plots_8}
# plotting a bar chart to visualize the distribution
ggplot(aes(x = quality.group), data = df) +
  geom_bar() + 
  scale_x_discrete()
```



### Structure of the dataset

The data set consists of 12 variables and 4898 observations. 
The individual data is either numerical or an integer.

### Main feature of interest in the dataset

Relation between the chemical properties of a wine and its quality.

### Other features of interest 

It will be interesting to see if the amount of alcohol in a wine has an effect on its density.
A general assumption is that alcohol has a lower density than water.

### Additional variables

I created bound.sulfur.dioxide by substracting free.sulfur.dioxide from total.sulfur.dioxide


# Bivariate Plots Section

```{r echo=FALSE, message=FALSE, Bivariate_Plots_1}
# creating a scatterplot for  fixed.acidity vs quality
ggplot(aes(x=fixed.acidity, y=quality), data = df) +
  geom_jitter(alpha = 1/5)
```

Because the variable _quality_ is an integer and therefore is discreet the plot shows "bars" which allow us to assume that there is no correlation between _quality_ and _fixed.acidity_. If the bars would be offset and there would be no vertical "overlap" it would suggest that _fixed.acidity_ has an effect on the quality of wine. 

```{r echo=FALSE, message=FALSE, Bivariate_Plots_2}
# creating a scatterplot for residual.sugar vs quality
ggplot(aes(x=residual.sugar, y=quality), data = df) +
  geom_jitter(alpha = 1/5)
```

More _residual.sugar_ does not imply a better quality of wine and vice-versa. The darker spots at each bar suggest that the majority of wines do not have much residual.sugar in them.

```{r echo=FALSE, message=FALSE, Bivariate_Plots_3}
# creating a scatterplot for alcohol vs quality
ggplot(aes(x=alcohol, y=quality), data = df) +
  geom_jitter(alpha = 1/5)
```

Wines with a quality of 5 and 6 seem to have alcohol levels which are lower than of those wines that have quality of over 6. Especially wines with a rating of 5 seem to have less alcohol. Wines with a quality of 7 tend to have alcohol levels above 10 whereas the lower qualities tend to mostly have alcohol levels starting at around 9. 

```{r echo=FALSE, message=FALSE, Bivariate_Plots_4}
# creating a scatterplot for volatile.acidity vs quality
ggplot(aes(x=volatile.acidity, y=quality), data = df) +
  geom_jitter(alpha = 1/5)
```

The _volatile.acidity_ of wines seem to have an effect on the quality. The lower bar suggests that wines with a quality of around 5 have more _volatile.acidity_ than wines with a quality of 6 or even 7.

```{r echo=FALSE, message=FALSE, Bivariate_Plots_5}
# creating a scatterplot for pH vs quality
ggplot(aes(x=pH, y=quality), data = df) +
  geom_jitter(alpha = 1/5)
```

The _ph_ factor of a wines seems to have no effect on the quality (almost parallel bars, no offset).

```{r echo=FALSE, message=FALSE, Bivariate_Plots_6}
# creating a scatterplot for density vs alcohol
ggplot(aes(y=alcohol, x=density), data = df) +
  geom_jitter(alpha = 1/5)
```

The plot shows that a high amount of alcohol leeds to a lower density and vice-versa.

```{r echo=FALSE, message=FALSE, Bivariate_Plots_7}
# creating a scatterplot for residual.sugar vs alcohol
ggplot(aes(y=alcohol, x=residual.sugar), data = df) +
  geom_jitter(alpha = 1/5)
```

The dark "border" suggests that a minimum amount of _residual.sugar_ is present in every wine at every alcohol level. The slope indicates that higher alcohol levels tend to have less _residual.sugar_ than their low level counterparts.

```{r echo=FALSE, message=FALSE, Bivariate_Plots_8}
# creating a scatterplot for pH vs fixed.acidity
ggplot(aes(y=fixed.acidity, x=pH), data = df) +
  geom_jitter(alpha = 1/5)
```

# Bivariate Analysis - Part 1

As per the previous plots it is challenging to point to any variable that has a fundamental effect on the quality of a wine. Unfortunately the data set lacks wines with a quality lower than 6 and more than 6. The summary() for quality shows that 50 percent of all wines are between grade 5 and 6.

```{r echo=FALSE, message=FALSE, Bivariate_Plots_add}
# getting a summary for the variable quality
summary(df$quality)
```

An interesting observation was that _residual.sugar_ stayed below 20 and therefore all wines in the data set were not sweet. 
It was furthermore interesting to see that there is a connection between the density of a wine and its alcohol content. This confirms whats meant to be common sense.
With alcohol being the outcome of the fermentation process it was interesting to see that remains of the sugar (which is used in the process) can be found in every wine. The chemical process seems to not completely turn sugar into alcohol. 


# Bivariate Analysis - Part 2 

A better understanding can maybe achieved by pairing the variables and calculation correlation coefficients.

## Correlation matrix

Because ggpairs produces a plot that is not useful I decided to use corrtable and corrplot().

### cortable
```{r echo=FALSE, message=FALSE, Multivariate_Plots_1}
cortable <- round(cor(subset(df, select=c("fixed.acidity", "volatile.acidity", 
                                          "citric.acid", "residual.sugar", 
                                          "chlorides", "free.sulfur.dioxide", 
                                          "total.sulfur.dioxide", "density", 
                                          "pH", "sulphates", "alcohol", 
                                          "quality"))), 2)
cortable
```

### corrplot

```{r echo=FALSE, message=FALSE, Multivariate_Plots_2}
# plotting a correlation matrix in the upper area and without the diagonal line
corrplot(cortable, type = "upper", diag = FALSE)
```

### Summary of cortable and corrplot

The table and the plot shows strong correlation (< -0.7, > +0.7) between:

- residual.sugar and density (+0.84)
- density and alcohol (-0.78)


The table and the plot shows moderate correlation (< -0.5, > +0.5) between:

- free.sulfur.dioxide and total.sulfur.dioxide (+0.62)
- total.sulfur.dioxide and density (+0.53)


# Multivariate Plots Section

### residual.sugar vs density for different qualities of wine

```{r echo=FALSE, message=FALSE, Multivariate_Plots_3}
# creating a scatterplot for residual.sugar vs density for the three quality
# groups including a conditional mean for every group
# limiting the axes by using quantile() and leaving out some extreme values
ggplot(aes(x = residual.sugar, y = density, 
           colour = quality.group), data = df) +
  geom_jitter(alpha = 1/4) +
  coord_cartesian( xlim = c(0, quantile(df$residual.sugar, 0.99)),
    ylim = c(quantile(df$density, 0.01), quantile(df$density, 0.99))) +
  geom_smooth(method = "loess")
```

By restricting the limits we generate a plot which is easier to interpret. The smoothed conditonal means allow us to conclude that wines of high quality have a low density and furthermore have less residual.sugar than wines of medium or low quality.

### alcohol vs density for different qualities of wine

```{r echo=FALSE, message=FALSE, Multivariate_Plots_4}
# creating a scatterplot for alcohol vs density for the three quality
# groups including a conditional mean for every group
# limiting the axes by using quantile() and leaving out some extreme values
ggplot(aes(x = alcohol, y = density, colour = quality.group), data = df) +
  geom_jitter(alpha = 1/4) +
  coord_cartesian(ylim = c(quantile(df$density, 0.01), 
                           quantile(df$density, 0.99))) +
  geom_smooth(method = "loess")
```

I cannot make out a difference between the three quality.groups as the smoothed conditional means are quite similar. However the plot shows that wines with a higher quality tend to have more alcohol in them (more blue points in the lower right area of the plot).


### free.sulfur.dioxide and total.sulfur.dioxide for different qualities of wine

```{r echo=FALSE, message=FALSE, Multivariate_Plots_5}
# creating a scatterplot for free.sulfur.dioxide vs total.sulfur.dioxide 
# for the three quality groups including a conditional mean for every group
# limiting the axes by using quantile() and leaving out some extreme values
ggplot(aes(x = free.sulfur.dioxide, y = total.sulfur.dioxide, 
           colour = quality.group), data = df) +
  geom_jitter(alpha = 1/4) +
  coord_cartesian(xlim = c(0, quantile(df$free.sulfur.dioxide, 0.99)), 
                  ylim = c(0, quantile(df$total.sulfur.dioxide, 0.99))) +
  geom_smooth(method = "loess")
```

Wines of high quality tend to have less amounts of free.sulfur.dioxide and total.sulfur.dioxide.

### total.sulfur.dioxide and density for different qualities of wine

```{r echo=FALSE, message=FALSE, Multivariate_Plots_6}
# creating a scatterplot for total.sulfur.dioxide vs density for the three 
# quality groups including a conditional mean for every group and with every
# group having another color
# limiting the axes by using quantile() and leaving out some extreme values
ggplot(aes(x = total.sulfur.dioxide, y = density, 
           colour = quality.group), data = df) +
  geom_jitter(alpha = 1/4) +
  coord_cartesian(xlim = c(0, 
                           quantile(df$total.sulfur.dioxide, 0.99)), 
                  ylim = c(quantile(df$density, 0.01), 
                           quantile(df$density, 0.99))) +
  geom_smooth(method = "loess")
```

In general the wines of medium quality seem to have a higher density and a higher amount of total.sulfur.dioxide. 
This does not apply to high quality wines which tend to be less dense.

# Multivariate Analysis

### Talk about some of the relationships you observed in this part of the \
investigation. Were there features that strengthened each other in terms of \
looking at your feature(s) of interest?

The most interesting relationships were those I was not aware of by doing a bivariate analysis because I focussed it on quality.

- The relationship between residual.sugar and density was quite interesting to see but on the other hand quite logical (generally speaking).
- The relationship between alcohol vs density was also quite predictable.
- The relationship between total.sulfur.dioxide and density was interesting.

### Were there any interesting or surprising interactions between features?

Almost all interactions were quite interesting because the pairs gave room for further assumptions.
High amounts of residual.sugar leed to a denser wine. A dense wine has a lower alcohol content than a wine which is less dense. Therefore one could conclude that 

------

# Final Plots and Summary

### Plot One
```{r echo=FALSE, message=FALSE, Plot_One}
# creating a scatterplot for alcohol vs residual.sugar for the three 
# quality groups with every group having another color 
ggplot(aes(y=alcohol, x=residual.sugar, color = quality.group), data = df) +
  geom_jitter(alpha = 1/4) +
  labs(title='alcohol vs residual.sugar') +
  xlab("residual.sugar in g/dm^3") +
  ylab("alcohol in % by volume")
```

### Description One

The plot shows the general relationship between residual.sugar and alcohol. It shows that the less residual.sugar is present, the higher is the amount of alcohol. Furthermore it shows that high quality wines have a lower amount of residual.sugar and therefore a higher amount of alcohol.

### Plot Two
```{r echo=FALSE, message=FALSE, Plot_Two}
# creating a scatterplot for alcohol vs density for the three 
# quality groups with every group having another color
ggplot(aes(y=alcohol, x=density, colour = quality.group), data = df) +
  geom_jitter(alpha = 1/4) +
  labs(title='alcohol vs density') +
  xlab("density in g/dm^3") +
  ylab("alcohol in % by volume")
```

### Description Two

This plot also shows the general relationship between density and alcohol. A high amount of alcohol lowers the density of a wine and vice-versa. Wines of high quality show higher amounts of alcohol and are therefore less dense than their counterparts of medium or low quality.


### Plot Three
```{r echo=FALSE, message=FALSE, Plot_Three}
# creating a scatterplot for residual.sugar vs density for the three 
# quality groups with every group having another color
# limiting the axes by using quantile() and leaving out some extreme values
ggplot(aes(x = residual.sugar, y = density, 
           colour = quality.group), data = df) +
  geom_jitter(alpha = 1/4) +
  coord_cartesian( xlim = c(0, quantile(df$residual.sugar, 0.99)),
    ylim = c(quantile(df$density, 0.01), quantile(df$density, 0.99))) +
  labs(title='residual.sugar vs density') +
  xlab("density in g/dm^3") +
  ylab("residual.sugar in g/dm^3")
```

### Description Three

This plot illustrates that a wine becomes more dense the more residual.sugar is present. It shows also that there seems to be a minimum amount of residual.sugar being present in every quality. This seems logical because of the chemical process that produces the alcohol in a wine. 


------

# Reflection

According to the dataset a wine is of high quality if it has a rather high amount of alcohol in it. With a high amount of alcohol generally comes a lower density. Not only does the amount of alcohol influence the density of wine but residual.sugar does it too. 

The investigation into the dataset provided an understanding of the relationship of some chemicals that play a role in the production of a wine. The strongest correlation to quality is alcohol so any future investigation should look into what has an effect on alcohol and therefore on the quality of a wine.

A rather large problem is the diversity of the dataset. As mentioned earlier almost 75% of the wines have a quality of 5 or 6. A wider spread in terms of quality would make the visual investigation a easier and would further provide a better insight into the data. Future analysis could include data about red wines and then investigate the relationships for every chemical of every type of wine and for both types combined.

# References

- general comments on the dataset: https://s3.amazonaws.com/udacity-hosted-downloads/ud651/wineQualityInfo.txt
- filter()-funcion: https://www.datanovia.com/en/lessons/subset-data-frame-rows-in-r/